---
title: 'Data Wrangling: Human Activity Recognition'
author: "Jonathan Freeland"
output: html_document
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(dplyr)
library(tidyr)
```
Wrangling performed on the [Samsung Galaxy S smartphone dataset](https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip) provided by UCI.
The following code will:

* Merge the training and test sets to create one data set.
* Extract columns containing mean and standard deviation for each measurement.
      * These column names contain "mean()" and "std()", respectively.
* Create variables ActivityLabel and ActivityName to label all observations via joining with the activity labels data set.
* Create a tidy data set with the average of each variable for each activity and subject.

***

Read in the activity labels and features.
```{r, cache=TRUE}
activities <- read_table("activity_labels.txt", col_names = FALSE)
features <- read_delim("features.txt", delim = " ", col_names = FALSE)
```

Create and execute a function to ingest the training and test data sets, including only mean and standard deviation columns.
```{r, cache=TRUE}
ingest_data <- function(setType) {
  # read in data
  subject <- read_table(paste0(setType,"/subject_",setType,".txt"), col_names = FALSE)
        x <- read_table(paste0(setType,"/X_",setType,".txt"), col_names = FALSE)
        y <- read_table(paste0(setType,"/y_",setType,".txt"), col_names = FALSE)
  # transpose features and assign as column names
  colnames(x) <- features[[2]]
  # select mean|std columns only
  x_select <- x %>% subset(select = grep("mean\\(\\)|std\\(\\)", names(x)))
  # add subject column
  x_select$SubjectId <- subject$X1
  # assign source type column
  x_select$SourceSet <- setType
  # add activity ID column
  x_select$ActivityLabel <- y$X1
  return(x_select)
}

testData <- ingest_data("test")
trainData <- ingest_data("train")
```

Combine the datasets.
```{r}
combinedData <- rbind(trainData, testData)
```

Join in the activity names and export the data set.
```{r}
colnames(activities) <- c("ActivityLabel", "ActivityName")
combinedDataJoined <- merge(combinedData, activities, by = "ActivityLabel")
write.csv(combinedDataJoined, file = "combinedDataJoined.csv")
```

Tidy up the data and export a data set of mean values by activity/subject.
```{r}
combinedDataTidy <- gather(combinedDataJoined, "Feature", "Value", 2:67)
groupedDataMeans <- combinedDataTidy %>% group_by_(c("ActivityLabel", "SubjectId", "ActivityName", "Feature")) %>% summarise(mean = mean(Value))
write.csv(groupedDataMeans, file = "groupedDataMeans.csv")
```


